---
title: "BBTV-Data_Exploration"
author: "Koissi Savi"
date: "2025-09-18"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries 

```{r, echo=FALSE, warning=FALSE}

library(tidyverse)
library(readxl)
library(stringdist)
library(stringr)
library(forcats)

#library(ggplot2)
library(skimr)       # for clean summary statistics
library(DataExplorer) # for quick EDA plots
library(sf)          # for mapping
library(leaflet)     # interactive map

library(janitor)

library(rnaturalearth)
library(rnaturalearthdata)

library(tidymodels)
library(vip)
library(xgboost)
library(gstat)

library(viridis)

library(brms)
library(posterior)


library(ggspatial)
library(terra)
library(ggrepel)
```

# Load data 

```{r, echo=FALSE, warning=FALSE}
# Extraction of metadata of BBTD sample collection

# Load data

data_cplt <-read_xlsx("/Users/koissi/Desktop/WAVE_project/Rapid_wave/BBTD/BBTD_transformed_06_2025_v2_complet.xlsx", sheet = "Plant_Dim")

meta<- read_xlsx("/Users/koissi/Desktop/WAVE_project/Rapid_wave/BBTD/BBTD_transformed_06_2025_v2_complet.xlsx", sheet = "Field_Dim")

cplt <- data_cplt %>%
  left_join(meta, by = "Unique_Field_ID")


```

# Data  cleaning

```{r, echo=FALSE, warning=FALSE}

data <- cplt %>%
  mutate(
    `Total Sucker Number` = as.numeric(`Total Sucker Number`),  # make sure it's numeric
    `Total Sucker Number` = if_else(
      is.na(`Total Sucker Number`),
      as.numeric(str_extract(Calculation, "(?<=-)[0-9]+(?=-[AS])")),
      `Total Sucker Number`
    )
  )%>%
  
  # Now rename as before
  rename(
    PlantNumber = `Plant Number`,
    TotalNsucker = "Total Sucker Number",
    SurveyedS = `Surveyed Sucker`,
    Disease_state = "Disease State",
    Severity = `Severity Score`,
    Aphid0wEL = `Aphid Count - Wingless (Emerging Leaf Sheath)`, 
    AphidEL   = `Aphid Count - Winged (Emerging Leaf Sheath)`,
    AphidB    = `Aphid Count - Winged (Base)`,
    Aphid0wB  = `Aphid Count - Wingless (Base)`, 
    AphidN    = `Aphid Count - Winged (Neck)`, 
    Aphid0wN  = `Aphid Count - Wingless (Neck)`,
    Location  = `Location name`, 
    AgroEco   = `Agro-ecology zone`, 
    Fieldsize = `Field size (m2)`,
    PlantationAge = `Musa field Age (Months)`,
    Distance  = `Distance traveled since the last field surveyed (km)`,
    NumberCountedField = `Number of Musa fields counted in the survey, after the last survey conducted`,
    Intercrops = `Intercrops within the field`,
    CropsSurrounding = `Crops around the field`,
    Management = `Managed or Unmanaged`,
    TypeFarming = `Production type`, 
    Latitude  = `_Automatic GPS Coordinates_latitude`,
    Longitude = `_Automatic GPS Coordinates_longitude`, 
    Altitude  = `_Automatic GPS Coordinates_altitude`, 
    CoordPrecision = `_Automatic GPS Coordinates_precision`
  ) %>%
  
  # Drop unwanted variables (including Calculation)
  dplyr::select(-c(
    "Calculation", "Picture", "start", "end", "Country", "YEAR OF COLLECTION", 
    "Field ID", "Surveyor Name", "Data Collector's Initials", "Enter a date and time", 
    "Country.1", "Manual coordinates/Latitude/Latitude hemisphere", 
    "Manual coordinates/Latitude/Degrees - latitude", 
    "Manual coordinates/Longitude/Longitude hemisphere", 
    "Manual coordinates/Longitude/Degrees - longitude", 
    "Manual coordinates/Longitude/Altitude(m)", 
    `Sample Collected`, `Unique Plant ID`,`Automatic GPS Coordinates`
  )) %>% 
  mutate( TotalAphid = Aphid0wEL + AphidEL + 
            AphidB + Aphid0wB + AphidN + Aphid0wN
    )

```

## Note: 
There are several qualitive variable that are transformed into a seemingly quantitative such as
* Severity                             
                              
 1 = no symptoms                      
 6 = bunchy top appearance            
 4 = chlorosis of leaf margin         
 2 = dark green streaks on leaf lamina
 5 = reduction in leaf size           
 3 = dark green streaks on the petiole

* Disease_state   
         
Asymptomatic  = 0  
Symptomatic  = 1    

```{r, echo=FALSE, warning=FALSE}
data_clean <- data %>%
  mutate(
    # Step 1: Clean categorical scores (extract only digits)
    across(c(Severity, CMV, BSV, BXW, Sigatoka, `Fusarium Wilt`, Management, Weediness),
           ~ str_extract(.x, "\\d+")),

    # Step 2: Recoding variables for consistency
    Symptoms = fct_recode(Symptoms,
      "No Symptoms" = " No symptoms",
      "No Symptoms" = "No symptoms Bunching",
      "No Symptoms" = "No symptoms Bunching Dot-dash symptoms on leaves Dot-dash symptoms on petioles or pseudostem",
      "No Symptoms" = "No symptoms Dot-dash symptoms on petioles or pseudostem"
    ),

    Disease_state = fct_recode(Disease_state,
      "0" = "Asymptomatic",
      "0" = "Asymptomatic = A",
      "1" = "Sympomatic",
      "1" = "Symptomatic",
      "1" = "Symptomatic = S"
    ),

    Intercrops = fct_recode(Intercrops,
      "1" = "Yes",
      "1" = "Manioc",
      "0" = "No",
      "0" = "Non"
    ), 
    Variety =fct_recode(Variety, 
                        "Banana plantain" = "banana plantain", 
                        "Sweet Banana" = "sweet banana"
      
    )
  ) %>%
  # Step 3: Convert cleaned numeric strings into numeric
  mutate(across(c(Severity, CMV, BSV, BXW, Sigatoka, `Fusarium Wilt`,
                  Management, Weediness), as.numeric)) %>%

  # Step 4: Harmonize crop names
  mutate(CropsSurrounding = str_replace_all(CropsSurrounding,
                                            c("Manihot esculenta" = "Manioc",
                                              "Maïs" = "Maize",
                                              "Mais" = "Maize", 
                                              "papayer" = "Papayer", 
                                              "palmier" = "Palmier "))) %>%

  # Step 5: Split multi-crop entries into separate rows
  mutate(CropsSurrounding = str_split(CropsSurrounding, pattern = ",\\s*|\\set\\s*")) %>%
  unnest(CropsSurrounding) %>%
  mutate(CropsSurrounding = str_trim(CropsSurrounding)) %>%

  # Step 6: One-hot encode crops with CS_ prefix
  mutate(value = 1) %>%
  pivot_wider(
    names_from = CropsSurrounding,
    values_from = value,
    values_fill = list(value = 0),
    names_prefix = "CS_"
  ) %>%

  # Step 7: Drop irrelevant dummy variables (noise categories)
  dplyr::select(-matches("CS_(NA|Non|N|No)$"))

data_clean <- data_clean %>%
  mutate(
    TypeFarming= if_else(
      Other == "Plantation/commercial Farm (intercrop)", 
      "Plantation/commercial Farm (intercrop)", 
      TypeFarming
    ),
    Other = if_else(
      Other == "Plantation/commercial Farm (intercrop)", 
      NA_character_, 
      Other
    )
  ) %>% 
  dplyr::select(-c(Other, Aphid0wEL, AphidEL, 
             AphidB, Aphid0wB, AphidN, Aphid0wN))

#data_clean %>% glimpse()
```

# Sanitation of the disease status

```{r, echo=FALSE, warning=FALSE}

data_clean <- data_clean %>%
  mutate(
    # safe numeric conversion (handles factors and characters)
    Disease_state = as.numeric(as.character(Disease_state)),
    Severity      = as.numeric(as.character(Severity)),

    # normalize symptoms text for reliable comparison
    Symptoms_clean = str_to_lower(str_trim(as.character(Symptoms))),

    # define evidence of disease: symptom not "no symptoms" OR severity not 1 (but only if severity is present)
    evidence_disease = ( !is.na(Symptoms_clean) & Symptoms_clean != "no symptoms" ) |
                       ( !is.na(Severity) & Severity != 1 ),

    # final decision rule
    Disease_state = case_when(
      Disease_state == 1 ~ 1,                                         # keep confirmed 1s
      (is.na(Disease_state) | Disease_state == 0) & evidence_disease ~ 1,  # upgrade to 1 when there is evidence
      TRUE ~ 0                                                         # otherwise mark 0
    )
  ) %>%
  dplyr::select(-Symptoms_clean, -evidence_disease)   # clean helper cols

# Symptoms
data_clean <- data_clean %>%
  mutate(
    # If Symptoms missing but Disease_state is 0, impute "No symptoms"
    Symptoms = if_else(
      is.na(Symptoms) & Disease_state == 0,
      "No symptoms",
      Symptoms
    ))

data_clean <- data_clean %>% 
  mutate(
    Severity = if_else(
      is.na(Severity) & Disease_state == 0,
      1,
      Severity
    )
  )

data_clean <- data_clean %>%
  mutate(
    SurveyedS = if_else(
      is.na(SurveyedS) & !is.na(TotalNsucker),
      TotalNsucker,
      SurveyedS
    )
  )

data_clean <- data_clean %>% 
  mutate(
    Disease_state = as.factor(Disease_state), 
    Severity = as.factor(Severity),
    Intercrops = as.factor(Intercrops),
    Management = as.factor( Management),
    Weediness = as.factor(Weediness),
    AgroEco = as.factor(AgroEco), 
    CS_Papayer = as.factor(CS_Papayer),
    CS_Palmier = as.factor(CS_Palmier),
    CS_manguier = as.factor(CS_manguier),
    CS_Amarante = as.factor(CS_Amarante),
    CS_Eucalyptus = as.factor(CS_Eucalyptus),
    CS_cocotiers = as.factor(CS_cocotiers),
    CS_Citronnelle = as.factor( CS_Citronnelle),
    CS_Ananas = as.factor(CS_Ananas),
    CS_Manioc = as.factor(CS_Manioc),
    CS_Vernonia = as.factor(CS_Vernonia),
    CS_basilicum = as.factor(CS_basilicum),
    CS_Maize = as.factor(CS_Maize),
    `CS_Canne à sucre` = as.factor(`CS_Canne à sucre`),
    CS_Basilic = as.factor(CS_Basilic),
    CS_vernonia = as.factor(CS_vernonia),
    `CS_grande morelle` = as.factor(`CS_grande morelle`),
    CS_Assissrè = as.factor(CS_Assissrè),
    CS_Igname = as.factor(CS_Igname),
    CS_Gombo = as.factor(CS_Gombo),
    CS_arachide = as.factor(CS_arachide),
    CS_haricot = as.factor(CS_haricot),
    CMV = as.factor(CMV),
    BSV = as.factor(BSV),
    BXW = as.factor(BXW),
    Sigatoka = as.factor(Sigatoka),
    `Fusarium Wilt` = as.factor(`Fusarium Wilt`)
    )
  
#data_clean %>% glimpse()

```



# Summary statistic 

```{r, echo=FALSE, warning=FALSE}

# 1. Summary statistics ----
# Separate numeric and categorical summaries
numeric_summary <- data_clean %>%
  dplyr::select(where(is.numeric)) %>%
  skim()

categorical_summary <- data_clean %>%
  dplyr::select(where(~is.character(.) | is.factor(.))) %>%
  skim()

# Display summaries
print(numeric_summary)
print(categorical_summary)

# If you want a cleaner export table

tab_num <- data_clean %>%
  dplyr::select(where(is.numeric)) %>%
  summary() %>%
  as.data.frame()

tab_cat <- data_clean %>%
  dplyr::select(where(~is.character(.) | is.factor(.))) %>%
  map_df(~ data.frame(var = deparse(substitute(.x)), 
                      levels = unique(.x), 
                      n = length(.x)))
# # 2. Quick graphs ----
# 
# 
# 
# # Frequency table as data.frame
# severity_df <- data_clean %>%
#   count(Severity)
# 
# # Bar plot
# ggplot(severity_df, aes(x = factor(Severity), y = n)) +
#   geom_col(fill = "steelblue") +
#   geom_text(aes(label = n), vjust = -0.5, size = 4) +
#   labs(
#     title = "Distribution of Severity Scores",
#     x = "Severity Score",
#     y = "Count"
#   ) +
#   theme_minimal()
# 
# 
# # Example: Disease state distribution
# ggplot(data_clean, aes(x = factor(Disease_state), 
#                        fill = factor(Disease_state))) +
#   geom_bar() +
#   theme_minimal() +
#   labs(title = "Distribution of Disease States")

# 3. Quick map ----


# 1. Convert your data into sf object
data_sf <- data_clean %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)

# 2. Get basemap for Benin (adjust country if needed)
benin <- ne_countries(scale = "medium", country = "Benin", returnclass = "sf")

# 3. Plot Disease_state on the map

ggplot() +
  # Country outline
  geom_sf(data = benin, fill = NA, color = "black", size = 0.6) +
  
  # Disease state points
  geom_sf(
    data = data_sf,
    aes(color = Disease_state),
    size = 2.5, alpha = 0.8
  ) +
  
  # Color scheme (color-blind safe, contrasting)
  scale_color_manual(
    values = c("0" = "#1b9e77", "1" = "#d95f02"),
    name = "Disease State",
    labels = c("0" = "Healthy", "1" = "Diseased")
  ) +
  
  # Add scale bar + north arrow
  annotation_scale(location = "bl", width_hint = 0.3, 
                   text_cex = 1.1, line_width = 0.8) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering,
                         height = unit(1.2, "cm"), width = unit(1.2, "cm")) +
  
  # Coordinate system
  coord_sf() +
  
  # Clean theme
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "gray85", size = 0.2),
    legend.position = "right",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 11),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  
  # Labels
  labs(
    title = "Spatial Distribution of Banana Bunchy Top Disease (BBTD)",
    subtitle = "Field survey, Benin (2025)",
    x = "Longitude", y = "Latitude"
  )


# Alternative 



# ---------------------------
# 1. Load admin level 2 shapefile
# ---------------------------
benin_lvl2 <- st_read("/Users/koissi/Downloads/ben/ben_admbnda_adm2_1m_salb_20190816.shp")

# ---------------------------
# 2. Collapse disease info at the field level
# ---------------------------
farm_status <- data_clean %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  group_by(Unique_Field_ID, Longitude, Latitude) %>%
  summarise(
    Diseased_any = ifelse(any(Disease_state == 1, na.rm = TRUE), 1, 0),
    .groups = "drop"
  )

# ---------------------------
# 3. Convert to sf points
# ---------------------------
data_sf <- st_as_sf(farm_status, coords = c("Longitude", "Latitude"),
                    crs = 4326, remove = FALSE)

# ---------------------------
# 4. Plot
# ---------------------------
ggplot() +
  # Admin boundaries
  geom_sf(data = benin_lvl2, fill = NA, color = "gray40", size = 0.5) +
  
  # Farm points
  geom_sf(
    data = data_sf,
    aes(color = factor(Diseased_any)),
    size = 2, alpha = 0.85
  ) +
  
  # Clear legend
  scale_color_manual(
    values = c("0" = "#1b9e77", "1" = "#d95f02"),
    name = "Field Status",
    labels = c("0" = "All Healthy", "1" = "At least 1 Diseased Plant")
  ) +
  
  # Add scale bar and north arrow
  annotation_scale(location = "bl", width_hint = 0.25,
                   text_cex = 1.1, line_width = 0.8) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering,
                         height = unit(1.2, "cm"), width = unit(1.2, "cm")) +
  
  coord_sf() +
  
  # Clean theme
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major = element_line(color = "gray85", size = 0.2),
    legend.position = "right",
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 13),
    axis.text = element_text(size = 11),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  
  # Labels
  labs(
    #title = "Field-Level Distribution of Banana Bunchy Top Disease (BBTD)",
    #subtitle = "Benin, 2025 — Admin Level 2 Boundaries",
    x = "Longitude", y = "Latitude"
  )




```

## Summary statistics

```{r,echo=FALSE, warning=FALSE}



# ---------------------------
# 2. Farms as sf points
# ---------------------------
farm_status <- data_clean %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  group_by(Unique_Field_ID, Longitude, Latitude) %>%
  summarise(
    Diseased_any = ifelse(any(Disease_state == 1, na.rm = TRUE), 1, 0),
    .groups = "drop"
  )

farm_sf <- st_as_sf(farm_status,
                    coords = c("Longitude", "Latitude"),
                    crs = 4326)

farm_sf <- st_transform(farm_sf, st_crs(benin_lvl2))

# ---------------------------
# 3. Spatial join: assign farms to admin 2 regions
# ---------------------------
farm_admin <- st_join(farm_sf, benin_lvl2, join = st_within)


farm_admin <- farm_admin %>%
  mutate(
    adm1_name = ifelse(is.na(adm1_name) & Unique_Field_ID == "F093_0",
                       "Oueme", adm1_name),
    adm2_name = ifelse(is.na(adm2_name) & Unique_Field_ID == "F093_0",
                       "Seme-Kpodji", adm2_name)
  )


# ---------------------------
# 4. Summary statistics per Admin 2 region
# ---------------------------
summary_stats <- farm_admin %>%
  st_drop_geometry() %>%
  group_by(adm1_name) %>%   # adjust column name depending on shapefile
  summarise(
    Total_fields   = n(),
    Infected_fields = sum(Diseased_any, na.rm = TRUE),
    Incidence      = mean(Diseased_any, na.rm = TRUE),
    SE              = sd(Diseased_any, na.rm = TRUE) / sqrt(n())#, 
    #Min             = min(Diseased_any, na.rm = TRUE),
    #Max             = max(Diseased_any, na.rm = TRUE)
  ) %>%
  arrange(desc(Incidence))

# ---------------------------
# 5. Display table
# ---------------------------
print(summary_stats)

```


With a better projection

```{r, echo=FALSE, warning=FALSE}
# Merge polygons by department (adm1_name)
benin_dept <- benin_lvl2 %>%
  group_by(adm1_name) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup()

# Compute centroid for each polygons
dept_labels <- benin_dept %>%
  st_centroid() %>%
  st_transform(crs = 4326) %>%  # ensure lat/lon CRS
  cbind(st_coordinates(.)) %>%  # get LONG/LAT columns
  as.data.frame() %>%
  dplyr::rename(Longitude = X, Latitude = Y)

# Map
ggplot() +
  # Department polygons with fill and legend
  geom_sf(data = benin_dept, aes(fill = adm1_name), color = "gray40", size = 0.5, alpha = 0.3) +
  
  # Farm points
  geom_sf(data = data_sf,
          aes(color = factor(Diseased_any)),
          size = 1.5, alpha = 0.85) +
  
  # Department labels (once per department)
  geom_text_repel(data = dept_labels,
                  aes(x = Longitude, y = Latitude, label = adm1_name),
                  size = 3, fontface = "bold") +
  
  # Legend for farm points
  scale_color_manual(
    values = c("0" = "#1b9e77", "1" = "#d95f02"),
    name = "Field Status",
    labels = c("0" = "All Healthy", "1" = "At least 1 Diseased Plant")
  ) +
  
  # Legend for department fill as squares
  scale_fill_discrete(name = "Department") +
  
  # Scale bar + north arrow
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering) +
  
  # Coordinate system
  coord_sf() +
  
  # Theme
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 11)
  )

### Alternative 
ggplot() +
  # Department polygons with fill and legend
  geom_sf(data = benin_dept, aes(fill = adm1_name), color = "gray40", size = 0.5, alpha = 0.3) +
  
  # Farm points
  geom_sf(data = data_sf,
          aes(color = factor(Diseased_any)),
          size = 1.5, alpha = 0.85) +
  
  # Legend for farm points
  scale_color_manual(
    values = c("0" = "#1b9e77", "1" = "#d95f02"),
    name = "Field Status",
    labels = c("0" = "All Healthy", "1" = "At least 1 Diseased Plant")
  ) +
  
  # Legend for department fill as squares
  scale_fill_discrete(name = "Department") +
  
  # Scale bar + north arrow
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering) +
  
  # Coordinate system
  coord_sf() +
  
  # Theme
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 11)
  )


```

# Extract climate data 

```{r, echo=FALSE, warning=FALSE}
library(terra)
library(sf)
library(dplyr)
library(readr)

# 1. Define folder containing climate rasters
climate_folder <- "/Volumes/KS_Onco/WhiteFliesProject/ClimateTop/"

# 2. List all raster files
climate_files <- list.files(climate_folder, full.names = TRUE,
                            pattern = "\\.(tif|nc)$", recursive = TRUE)

# 3. Load boundary for Benin (you already have benin_lvl2 or benin_dept)
# Just take the outer boundary (country level)
benin_boundary <- st_union(benin_lvl2) |> vect()

# 4. Crop each raster to Benin extent before resampling
terraOptions(tempdir = "/Volumes/KS_Onco/tmp")

climate_list <- lapply(climate_files, function(f) {
  r <- rast(f)
  r_crop <- crop(r, benin_boundary)        # crop to Benin extent
  mask(r_crop, benin_boundary)             # mask outside Benin
})

# 5. Choose a reference (first cropped raster)
ref <- climate_list[[1]]

# 6. Resample all to same resolution/extent
climate_resampled <- lapply(climate_list, function(r) {
  resample(r, ref, method = "bilinear")
})

# 7. Stack them
climate_stack <- rast(climate_resampled)

# 8. Prepare farm points as sf
farm_sf <- data_clean %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# Reproject farm points to match raster CRS
if (!st_crs(farm_sf) == crs(climate_stack)) {
  farm_sf <- st_transform(farm_sf, crs(climate_stack))
}

# 9. Extract climate values
climate_values <- terra::extract(climate_stack, vect(farm_sf)) %>%
  dplyr::select(-ID)

# 10. Merge extracted values with farm dataset
data_climate <- cbind(data_clean %>% filter(!is.na(Longitude)), climate_values)

# 11. Save merged dataset
write_csv(data_climate, "Merged_Climate_BBTD_Farms.csv")

head(data_climate)


```
Re-classify variables if you started with the new dataset

```{r, echo=FALSE, warning=FALSE}
data_climate <- read_csv("Merged_Climate_BBTD_Farms.csv")

# Identify all CS columns
cs_cols <- grep("^CS_", names(data_climate), value = TRUE)

data_climate <- data_climate %>%
  # Convert selected columns to factors
  mutate(across(c(Disease_state, Severity, CMV, BSV, BXW, Sigatoka, `Fusarium Wilt`, AgroEco, 
                  Intercrops, Management, Weediness, all_of(cs_cols)), as.factor)) %>% 
  dplyr::select(-c(Unique_Field_ID, PlantNumber, Symptoms, Severity,
            Departement, AgroEco, Commune, Location, TypeFarming  
            )) %>% 
  sf::st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

dat_cli <- data_climate %>%
  # Convert selected columns to factors
  mutate(across(c(Disease_state, Severity, CMV, BSV, BXW, Sigatoka, `Fusarium Wilt`, AgroEco, 
                  Intercrops, Management, Weediness, all_of(cs_cols)), as.factor)) %>% 
  dplyr::select(-c( Symptoms, Severity,
            Departement, AgroEco, Commune, Location, TypeFarming  
            )) %>% 
  sf::st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# # Clean and collapse CS columns
# data_climate_clean <- data_climate %>%
#   # Convert selected columns to factors
#   mutate(across(c(Disease_state, Severity, CMV, BSV, BXW, Sigatoka, `Fusarium Wilt`, AgroEco, 
#                   Intercrops, Management, Weediness, all_of(cs_cols)), as.factor)) %>%
#   
#   # Collapse CS columns into a single "CS" column
#   pivot_longer(cols = all_of(cs_cols), names_to = "CS_crop", values_to = "CS_value") %>%
#   mutate(CS_crop = sub("^CS_", "", CS_crop)) %>%
#   filter(CS_value == 1) %>%
#   select(-CS_value) %>%
#   
#   # Aggregate CS crops by other variables
#   group_by(Disease_state, Severity, CMV, BSV, BXW, Sigatoka, `Fusarium Wilt`, AgroEco,
#            Intercrops, Management, Weediness, Longitude, Latitude) %>%
#   summarise(CS = paste(CS_crop, collapse = ", "), .groups = "drop") %>%
#   
#   # Convert to sf object
#   sf::st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)


```

split the matrix 

```{r, echo=FALSE, warning=FALSE}

# Separate numeric and categorical predictors
num_vars <- data_climate %>% dplyr::select(where(is.numeric))
cat_vars <- data_climate %>% dplyr::select(where(is.character), where(is.factor))

# Remove numeric vars with >30% NA
num_vars <- num_vars %>% select(where(~ mean(is.na(.)) < 0.3))
cat_vars <- cat_vars%>% select(where(~ mean(is.na(.)) < 0.3))

#Remove near-zero variance predictors; a predictor variable in a dataset where almost all values are the same, or the distribution of values is highly unbalanced



library(recipes)
# Convert to plain data frame (drop sf class + geometry)
# Drop geometry first
dimm <- data_climate %>% 
  sf::st_drop_geometry()


rec <- recipe(Disease_state ~ ., data = dimm) %>%
  step_unknown(all_nominal_predictors(), new_level = "Unknown") %>%  # ✅ handle NA as "Unknown"
  step_dummy(all_nominal_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_corr(all_numeric_predictors(), threshold = 0.6) %>%
  step_lincomb(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())


# Now prep the recipe
prepped <- prep(rec, training = dimm)


# Bake it (apply preprocessing to the dataset)
final_data <- bake(prepped, new_data = dimm)

# Inspect the result
dim(final_data)
head(final_data)





```

Prepare dataset for modeling 

```{r, echo=FALSE, warning=FALSE}
library(dplyr)
library(caret)
library(xgboost)
library(pROC)

# Remove geometry before ML
ml_data <- final_data

# Ensure outcome is numeric 0/1
ml_data$Disease_state <- as.numeric(as.character(ml_data$Disease_state))

# Separate numeric and categorical again
num_features <- ml_data %>% select(where(is.numeric))
cat_features <- ml_data %>% select(where(is.factor) | where(is.character))


pre_proc_num <- preProcess(num_features, method = c("medianImpute"))
num_features <- predict(pre_proc_num, newdata = num_features)

mode_impute <- function(x) {
  x[is.na(x)] <- names(sort(table(x), decreasing = TRUE))[1]
  return(x)
}

cat_features <- cat_features %>% mutate(across(everything(), mode_impute))

ml_data <- bind_cols(num_features, cat_features)


```
# Modelling 

```{r, echo=FALSE, warning=FALSE}
set.seed(123)
train_index <- createDataPartition(ml_data$Disease_state, p = 0.8, list = FALSE)
train_data <- ml_data[train_index, ]
test_data <- ml_data[-train_index, ]

# xgboost model
# Convert Disease_state to numeric 0/1
train_data$Disease_state <- as.numeric(as.character(train_data$Disease_state))
train_data$Disease_state <- ifelse(train_data$Disease_state == 0, 0, 1)

test_data$Disease_state <- as.numeric(as.character(test_data$Disease_state))
test_data$Disease_state <- ifelse(test_data$Disease_state == 0, 0, 1)

# Now convert predictors to dummy variables
library(fastDummies)

# Training/test matrices for XGBoost
x_train <- as.matrix(train_data %>% select(-Disease_state))
y_train <- train_data$Disease_state

x_test <- as.matrix(test_data %>% select(-Disease_state))
y_test <- test_data$Disease_state

# XGBoost DMatrix
dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dtest  <- xgb.DMatrix(data = x_test, label = y_test)


params <- list(
  objective = "binary:logistic",
  eval_metric = "auc"
)

xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 200,
  watchlist = list(train = dtrain, eval = dtest),
  early_stopping_rounds = 10,
  verbose = 1
)

importance <- xgb.importance(model = xgb_model)
head(importance)  # top 20 important predictors

library(ggplot2)
library(dplyr)

# Select top 20 features by Gain
top20 <- importance %>%
  arrange(desc(Gain)) %>%
  slice_head(n = 20) %>%
  mutate(Feature = factor(Feature, levels = rev(Feature)))  # for horizontal ordering

library(sensitivity)

# Make sure X is a data.frame
X <- as.data.frame(x_train)
Y <- as.numeric(y_train)   # must be numeric vector (0/1)

# Run PRCC with bootstrapping
prcc_res <- pcc(X, Y, rank = TRUE, nboot = 100)

# Extract results
prcc_df <- as.data.frame(prcc_res$PRCC)
prcc_df$Feature <- rownames(prcc_df)
colnames(prcc_df)
colnames(prcc_df)[1] <- "PRCC"
top20 <- importance %>%
    arrange(desc(Gain)) %>%
    slice_head(n = 20) %>%
    left_join(prcc_df, by = "Feature") %>%
    mutate(Feature = factor(Feature, levels = rev(Feature)))

ggplot(top20, aes(x = Feature, y = Gain, fill = PRCC)) +
    geom_col() +
    scale_fill_gradient2(
        low = "steelblue",
        mid = "white",
        high = "firebrick",
        midpoint = 0,
        name = "PRCC\n(Direction)"
    ) +
    coord_flip() +
    labs(
        #title = "Top 20 Predictors of Disease State",
        x = "Predictors",
        y = "Importance (Gain)"
    ) +
    theme_minimal(base_size = 16) +
    theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "right",
        panel.grid.major.y = element_blank()
    )


# Get predictions on test set
pred <- predict(xgb_model, x_test)
roc_obj <- roc(y_test, pred)

ggroc(roc_obj, color = "firebrick", size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  labs(
    title = "Cross-validated ROC Curve",
    subtitle = paste("AUC =", round(auc(roc_obj), 3)),
    x = "1 - Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )



```


# What can explain the occurence of the BBTD?

```{r, echo=FALSE, warning=FALSE}

set.seed(123)  # reproducibility
data_ml <- data_climate %>%
  dplyr::select(-c(Unique_Field_ID, PlantNumber, 
            Departement, 
            Commune, 
            Location, 
            AgroEco, 
            TypeFarming, 
            CoordPrecision,
            Latitude,
            Longitude,
            Severity, 
            Symptoms)) %>% 
  na.omit()   # simple handling of missing values for now

split <- initial_split(data_ml, prop = 0.8, strata = Disease_state)
train_data <- training(split)
test_data  <- testing(split)

rec <- recipe(Disease_state ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%   # convert factors to dummies
  step_zv(all_predictors()) %>%              # remove zero variance predictors
  step_normalize(all_numeric_predictors())   # scale numeric features

# rf_model <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) %>%
#   set_engine("ranger") %>%
#   set_mode("classification")

rf_model <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")


wf <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rec)

set.seed(123)
cv_folds <- vfold_cv(train_data, v = 5, strata = Disease_state)

tuned <- tune_grid(
  wf,
  resamples = cv_folds,
  grid = 10,
  metrics = metric_set(roc_auc, accuracy, sens, spec)
)

best_params <- show_best(tuned, metric = "roc_auc", n = 1)

final_rf <- finalize_workflow(wf, best_params)

final_fit <- last_fit(final_rf, split)

collect_metrics(final_fit)

# Get predictions
preds <- collect_predictions(final_fit)
conf_mat <- conf_mat(preds, truth = Disease_state, estimate = .pred_class)
conf_mat



final_model <- extract_fit_parsnip(final_fit)
vip(final_model, num_features = 25)

```

With another framework more powerfull. Let us use the xgboost

## What is XGBoost?

XGBoost (Extreme Gradient Boosting) is a powerful machine learning algorithm that uses boosted decision trees.

* Compared to Random Forest:

RF builds many independent trees (bagging).

XGBoost builds trees sequentially, where each new tree corrects errors of the previous ones (boosting).

It’s fast, handles missing values, works well with tabular data, and often outperforms RF in classification problems like your disease-state prediction.


```{r, echo=FALSE, warning=FALSE}
xgb_model <- boost_tree(
  trees = 1000,        # number of trees
  tree_depth = tune(), # depth of each tree
  learn_rate = tune(), # learning rate (shrinkage)
  loss_reduction = tune(), # min split gain
  mtry = tune(),       # predictors per split
  sample_size = tune() # subsample fraction
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

wf_xgb <- workflow() %>%
  add_model(xgb_model) %>%
  add_recipe(rec)   # same recipe you already built

set.seed(123)
tuned_xgb <- tune_grid(
  wf_xgb,
  resamples = cv_folds,
  grid = 20,  # more grid points since xgboost has more params
  metrics = metric_set(roc_auc, accuracy, sens, spec)
)


best_xgb <- select_best(tuned_xgb, metric = "roc_auc")
final_wf_xgb <- finalize_workflow(wf_xgb, best_xgb)
final_fit_xgb <- last_fit(final_wf_xgb, split = split)
collect_metrics(final_fit_xgb)


# Extract the trained xgboost model
final_model_xgb <- extract_fit_parsnip(final_fit_xgb$.workflow[[1]])

# Plot top 15 most important features
vip(final_model_xgb, num_features = 15) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::labs(
    title = "Top Predictors of Disease State (XGBoost)",
    x = "Importance",
    y = "Feature"
  )


```

# Compute the apparent incidence vs true incidence

## Prep for data analysis

```{r, echo=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(recipes)

data_climate <- read_csv("Merged_Climate_BBTD_Farms.csv")

# Identify CS columns
cs_cols <- grep("^CS_", names(data_climate), value = TRUE)

# Convert to factors and prepare spatial object
dat_cli <- data_climate %>%
  mutate(across(c(Disease_state, Severity, CMV, BSV, BXW, Sigatoka, 
                  `Fusarium Wilt`, AgroEco, Intercrops, Management, 
                  Weediness, all_of(cs_cols)), as.factor)) %>%
  sf::st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  dplyr::select(-c(AgroEco, Symptoms, Severity))

dat_cli <- dat_cli %>% 
  dplyr::select(c(Unique_Field_ID, PlantNumber, TotalNsucker,
           SurveyedS, Disease_state,
           Departement, Fieldsize, 
           PlantationAge, Distance,
           NumberCountedField, Intercrops,
           Management,Weediness, CoordPrecision,  
           TotalAphid,CS_Maize,
           ORCDRC_M_sl1_250m_ll_organicMatter,
           wc2.1_30s_tmax_09, wc2.1_30s_wind_04,
           geometry 
           ))
# Drop geometry before recipe processing
dimm <- dat_cli %>% sf::st_drop_geometry()

final_data <-  dimm

# Inspect results
dim(final_data)
names(final_data)

covs <-c("TotalNsucker", "Fieldsize", "PlantationAge",  
         "Distance",  "NumberCountedField")                      
           

```
## Computing the observed incidence

```{r, echo=FALSE, warning=FALSE}


#### Or this with the geocoordinate



# Plot the prevalence map
library(tidyverse)
library(sf)
library(ggplot2)
library(ggspatial) # for scale bar and north arrow

# Extract coordinates from geometry
dat_cli_coords <- dat_cli %>%
  mutate(Longitude = st_coordinates(.)[,1],
         Latitude = st_coordinates(.)[,2]) %>%
  st_drop_geometry()

# Compute observed prevalence per farm
prevalence_per_farm <- dat_cli_coords %>%
  group_by(Unique_Field_ID, Longitude, Latitude) %>%
  summarise(
    diseased_plants = sum(Disease_state == 1, na.rm = TRUE),
    total_surveyed = sum(SurveyedS, na.rm = TRUE),
    prevalence = 100 * diseased_plants / total_surveyed,
    .groups = "drop"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# Read Benin shapefile
benin_lvl2 <- st_read("/Users/koissi/Downloads/ben/ben_admbnda_adm2_1m_salb_20190816.shp")

# Plot map with north arrow and scale bar
ggplot() +
  geom_sf(data = benin_lvl2, fill = "grey95", color = "grey70") +
  geom_sf(data = prevalence_per_farm, aes(color = prevalence), 
          size = 1.5, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma", name = "Incidence (%)") +
  annotation_scale(
    location = "bl", # bottom left
    width_hint = 0.3,
    bar_cols = c("grey80", "white")
  ) +
  annotation_north_arrow(
    location = "tl", # top left
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  theme_minimal() +
  labs(
    #title = "Observed BBTD Prevalence per Farm in Benin",
    #subtitle = "Calculated from field survey data",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "right"
  )


```
## Compute true prevalence 

```{r, echo=FALSE, warning=FALSE}
# Install if not already installed
# install.packages(c("brms", "tidyverse", "sf", "ggplot2", "ggspatial", "patchwork", "viridis"))

library(brms)
library(tidyverse)
library(sf)
library(ggplot2)
library(ggspatial)
library(patchwork)
library(viridis)

# Sensitivity and specificity priors (expert-informed)
Se_prior <- 0.7   # prior mean (sensitivity)
Sp_prior <- 0.9   # prior mean (specificity)
Se_sd <- 0.1
Sp_sd <- 0.05

# Compute observed prevalence
prevalence_per_farm <- dat_cli %>%
  mutate(Longitude = st_coordinates(.)[,1],
         Latitude = st_coordinates(.)[,2]) %>%
  st_drop_geometry() %>%
  group_by(Unique_Field_ID, Longitude, Latitude) %>%
  summarise(
    diseased_plants = sum(Disease_state == 1, na.rm = TRUE),
    total_surveyed = sum(SurveyedS, na.rm = TRUE),
    prevalence = 100 * diseased_plants / total_surveyed,
    .groups = "drop"
  )

# Fit Bayesian measurement-error model
# Latent p_i = true prevalence
# Observed X_i ~ Binomial(N_i, Se*p_i + (1 - Sp)*(1 - p_i))
# Install if needed
# install.packages(c("cmdstanr","posterior","bayesplot","rstan","sf","tidyverse","ggspatial","patchwork","viridis"))
# cmdstanr requires CmdStan installed; see cmdstanr docs.


library(posterior)
library(bayesplot)
library(patchwork)
library(viridis)

# Convert your plausible Se/Sp ranges to Beta priors (match central 95% to ranges)

# helper: find beta params matching two quantiles
beta_from_quantiles <- function(q_lo, q_hi, p_lo = 0.025, p_hi = 0.975){
  # optimize alpha,beta to match quantiles
  fn <- function(par){
    a <- par[1]; b <- par[2]
    if(a <= 0 || b <= 0) return(1e6)
    s <- (qbeta(p_lo, a, b) - q_lo)^2 + (qbeta(p_hi, a, b) - q_hi)^2
    return(s)
  }
  opt <- stats::optim(c(2,2), fn, method = "L-BFGS-B", lower = c(1e-6,1e-6))
  return(list(alpha = opt$par[1], beta = opt$par[2], value = opt$value))
}

# your plausible ranges
Se_range <- c(0.40, 0.90)
Sp_range <- c(0.80, 0.99)

Se_beta <- beta_from_quantiles(Se_range[1], Se_range[2])
Sp_beta <- beta_from_quantiles(Sp_range[1], Sp_range[2])

Se_beta; Sp_beta
# Use Se_beta$alpha, Se_beta$beta, etc.


```

## Prepare data for Stan
```{r}
# use the 'final_data' (glimpse provided earlier)
df <- final_data  # change if your object name differs
df <- df %>% mutate(disease_obs = as.integer(as.character(Disease_state))) 
# Ensure Disease_state is 0/1 numeric
# Step 1: Keep only covariates + outcome
df_stan <- df %>%
  dplyr::select(Unique_Field_ID, disease_obs, Departement, all_of(covs)) %>%
  filter(!is.na(disease_obs)) %>%
  mutate(across(all_of(covs), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))


# Step 2: Convert categorical variables to numeric dummy variables
X <- model.matrix(~ .,
                  data = df_stan %>% 
                    dplyr::select(all_of(covs))) %>%
     as.matrix()

stan_data$K <- length(unique(df_stan$field_idx))
# Step 3: Field index
field_ids <- df_stan %>%
  distinct(Unique_Field_ID) %>%
  arrange(Unique_Field_ID)

df_stan <- df_stan %>%
  left_join(field_ids %>% mutate(field_idx = row_number()),
            by = "Unique_Field_ID")

# Step 4: Stan data list
#set.seed(123)
stan_data <- list(
  N = nrow(df_stan),
  P = ncol(X),
  K = length(unique(df_stan$field_idx)),
  field = df_stan$field_idx,
  X = X,
  y = df_stan$disease_obs,
  Se_a = Se_beta$alpha,
  Se_b = Se_beta$beta,
  Sp_a = Sp_beta$alpha,
  Sp_b = Sp_beta$beta
)


# Add Beta prior params for Se/Sp computed earlier:
stan_data$Se_a <- Se_beta$alpha
stan_data$Se_b <- Se_beta$beta
stan_data$Sp_a <- Sp_beta$alpha
stan_data$Sp_b <- Sp_beta$beta
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()

library(cmdstanr)

mod <- cmdstan_model("~/Desktop/WAVE_project/Theory/CassavaMap/misclass_hier_logit.stan")

```
### Compilling model

```{r}

fit <- mod$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123,
  save_cmdstan_config = TRUE
)




# Extract draws as array
posterior_array <- fit$draws()  # returns draws_array

# Check parameter names
dimnames(posterior_array)$variable

# Suppose you want 'p[1]', 'p[2]', ..., 'p[N]'
library(posterior)

posterior_df <- as_draws_df(posterior_array)

# Select only p[i] columns
p_cols <- grep("^p\\[", names(posterior_df), value = TRUE)
posterior_mat <- as.matrix(posterior_df[, p_cols])
dim(posterior_mat)  # should be n_samples x N

dat_cli_row_coords <- dat_cli %>%
  mutate(Longitude = st_coordinates(.)[,1],
         Latitude = st_coordinates(.)[,2]) %>%
  st_drop_geometry()


posterior_long <- as.data.frame(posterior_mat) %>%
    pivot_longer(
        cols = everything(),
        names_to = "field_index",
        values_to = "p"
    ) %>%
    mutate(
        field_index = as.integer(gsub("p\\[|\\]", "", field_index))
    ) %>%
    left_join(
        dat_cli_row_coords %>%
            mutate(field_index = row_number()) %>%
            dplyr::select(field_index, Departement, Unique_Field_ID, Longitude, Latitude),
        by = "field_index"
    ) %>%
    group_by(Unique_Field_ID) %>%
    summarise(
        true_prev_median = median(p) * 100,
        true_prev_lower = quantile(p, 0.025) * 100,
        true_prev_upper = quantile(p, 0.975) * 100,
        Longitude = first(Longitude),
        Latitude = first(Latitude),
        .groups = "drop"
    ) %>%
    st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# Ensure data CRS matches Benin shapefile
posterior_long_sf <- st_transform(posterior_long, st_crs(benin_lvl2))
#library(sf)

# Make sure Longitude and Latitude exist in the data
prevalence_per_farm <- prevalence_per_farm %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# Now you can safely transform the CRS to match benin_lvl2
prevalence_per_farm <- st_transform(prevalence_per_farm, st_crs(benin_lvl2))



```
### True incidence

```{r}
library(sf)
library(dplyr)

# Convert observed prevalence to sf
prevalence_per_farm_sf <- st_as_sf(
  prevalence_per_farm,
  coords = c("Longitude", "Latitude"),
  crs = 4326 # WGS84 (standard for GPS)
)

# Make sure posterior_long is also sf
# posterior_long_sf <- st_as_sf(
#   posterior_df,
#   coords = c("Longitude", "Latitude"),
#   crs = 4326
# )



library(ggplot2)
library(sf)
library(ggspatial)
library(patchwork)

# --- Define shared color scale limits ---
common_limits <- c(
  0,
  max(
    c(prevalence_per_farm_sf$prevalence,
      posterior_long_sf$true_prev_median),
    na.rm = TRUE
  )
)

# --- A. Observed prevalence map ---


# --- Define common color limits ---
common_limits <- c(
  0,
  max(
    c(prevalence_per_farm_sf$prevalence,
      posterior_long_sf$true_prev_median),
    na.rm = TRUE
  )
)
##------ Summary stats
# Ensure CRS alignment
posterior_long_sf <- st_transform(posterior_long, st_crs(benin_lvl2))

# Spatial join: assign each posterior point to its Department (Admin 1)
posterior_admin <- st_join(posterior_long_sf, benin_lvl2, join = st_within)

# Summarize posterior-based incidence per Department (adm1_name)
posterior_summary <- posterior_admin %>%
  st_drop_geometry() %>%
  group_by(adm1_name) %>%
  summarise(
    True_prev_median = mean(true_prev_median, na.rm = TRUE),
    True_prev_lower = mean(true_prev_lower, na.rm = TRUE),
    True_prev_upper = mean(true_prev_upper, na.rm = TRUE),
    N_fields = n()
  ) %>%
  arrange(desc(True_prev_median))

# View results
posterior_summary


# --- A. Observed prevalence map ---
p_obs <- ggplot() +
  geom_sf(data = benin_lvl2, fill = "grey95", color = "grey70") +
  geom_sf(data = prevalence_per_farm_sf,
          aes(fill = prevalence),
          size = 2.5, shape = 21, color = "black", alpha = 0.9, stroke = 0.15) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "BBTD Incidence (%)",
    limits = common_limits
  ) +
  labs(title = "A. Observed BBTD Incidence") +
  annotation_scale(
    location = "bl",
    width_hint = 0.4,   # wider for 100 km
    bar_cols = c("grey80", "white"),
    text_cex = 0.9,
    line_width = 0.8
  ) +
  annotation_north_arrow(
    location = "tl",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

# --- B. True prevalence map ---
p_true <- ggplot() +
  geom_sf(data = benin_lvl2, fill = "grey95", color = "grey70") +
  geom_sf(data = posterior_long_sf,
          aes(fill = true_prev_median),
          size = 2.5, shape = 21, color = "black", alpha = 0.9, stroke = 0.15) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "BBTD Incidence (%)",
    limits = common_limits
  ) +
  labs(title = "B. Estimated True BBTD Incidence (95% CI)") +
  annotation_scale(
    location = "bl",
    width_hint = 0.4,
    bar_cols = c("grey80", "white"),
    text_cex = 0.9,
    line_width = 0.8
  ) +
  annotation_north_arrow(
    location = "tl",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )



# --- Combine side-by-side ---
p_obs | p_true

######------- Stat at the Department level
data_list <- list(
  N = nrow(summary_stats),
  y = ifelse(is.na(summary_stats$Infected_fields), 0, summary_stats$Infected_fields),
  n = ifelse(is.na(summary_stats$Total_fields), 0, summary_stats$Total_fields),
  Se_a = Se_beta$alpha,
  Se_b = Se_beta$beta,
  Sp_a = Sp_beta$alpha,
  Sp_b = Sp_beta$beta
)

mod <- cmdstan_model("model.stan")
fit <- mod$sample(
  data = data_list,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123
)
departments <- unique(summary_stats$adm1_name)

# Extract summary for p_true
posterior_summary_admin <- fit$summary("p_true") %>%
  as.data.frame() %>%
  mutate(
    Departement = summary_stats$adm1_name,  # make sure the order matches
    Posterior_mean_incidence = mean*100,
    Posterior_SE = sd,
    Posterior_Lower = q5,
    Posterior_Upper = q95
  ) %>%
  select(Departement, Posterior_mean_incidence, Posterior_SE, Posterior_Lower, Posterior_Upper)

posterior_summary_admin


# Join observed and Bayesian summaries
summary_compare <- summary_stats %>%
  left_join(
    posterior_summary_admin,
    by = c("adm1_name" = "Departement")
  ) %>%
  mutate(
    Bias = Posterior_mean_incidence - Observed_incidence
  ) %>%
  select(
    adm1_name, Total_fields, Infected_fields, Observed_incidence,
    Posterior_mean_incidence, Posterior_SE, Posterior_Lower, Posterior_Upper,
    Bias
  )

summary_compare



```


## Wavefront detection

```{r, echo=FALSE, warning=FALSE}
data2 <- read_csv("/Users/koissi/Downloads/banana_wave_2018.csv")

# 1️⃣ Rename columns in data2 to match those in final_data
data2_renamed <- data2 %>%
  rename(
    Unique_Field_ID = id,
    Latitude = lat,
    Longitude = lon,
    Departement = NAME_1,
    Commune = NAME_2,
    Disease_state = BBTD_inc
  ) %>% 
  select(-c(NAME_3, country_name, adm0_name,
            country_iso3, crop, date_created_unixtimestamp , 
            pest, source,
            date_created, date_accessed, date_created_unixtimestamp)) %>% 
  relocate( Disease_state, year, .after = year)

# 2️⃣ Add the year column (since final_data is from 2025)
dat <- data_climate %>%
  mutate(year = 2025) %>% 
  dplyr::select(Unique_Field_ID,Latitude,
         Longitude, Departement,
         Commune, Disease_state, year)

# 3️⃣ Join datasets on Unique_Field_ID
merged_data <- dat %>%
  rbind(data2_renamed)

## Compute true incidence per polygon

#Define the sensitivity and specificity 

# --- Priors (expert-informed) ---
Se_range <- c(0.40, 0.90)
Sp_range <- c(0.80, 0.99)

# Beta function to match quantiles
beta_from_quantiles <- function(q_lo, q_hi, p_lo = 0.025, p_hi = 0.975){
  fn <- function(par){
    a <- par[1]; b <- par[2]
    if(a <= 0 || b <= 0) return(1e6)
    s <- (qbeta(p_lo, a, b) - q_lo)^2 + (qbeta(p_hi, a, b) - q_hi)^2
    return(s)
  }
  opt <- stats::optim(c(2, 2), fn, method = "L-BFGS-B", lower = c(1e-6, 1e-6))
  list(alpha = opt$par[1], beta = opt$par[2])
}

Se_beta <- beta_from_quantiles(Se_range[1], Se_range[2])
Sp_beta <- beta_from_quantiles(Sp_range[1], Sp_range[2])


# Convert merged_data to sf object
merged_sf <- merged_data %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

benin_lvl2 <- benin_lvl2 %>% 
  rename(Commune = adm2_name,
         Departement = adm1_name)


# Spatial join: assign Commune where missing
merged_filled <- merged_sf %>%
  st_join(benin_lvl2 %>% select(Commune, Departement), join = st_intersects) %>%
  mutate(
    Commune = ifelse(is.na(Commune.x), Commune.y, Commune.x),
    Departement = ifelse(is.na(Departement.x), Departement.y, Departement.x)
  ) %>%
  select(-Commune.x, -Commune.y, -Departement.x, -Departement.y)

## Extract Coordinate 
merged_filled_coords <- merged_filled %>%
  mutate(
    Longitude = st_coordinates(.)[, 1],
    Latitude = st_coordinates(.)[, 2],
    year = ifelse(is.na(year), 2025, year), 
    disease_obs = as.integer(Disease_state)
  ) %>%
  st_drop_geometry()




# Create a unique field index
field_ids <- merged_filled_coords %>%
  distinct(Unique_Field_ID, year, Commune, Departement, Latitude, Longitude) %>%
  arrange(Unique_Field_ID, year, Commune, Departement,) %>%
  mutate(field_idx = row_number())


dft_stan <- merged_filled_coords %>%
  left_join(field_ids, by = "Unique_Field_ID")

# Create list of stan
stan_data <- list(
  N = nrow(dft_stan),
  K = length(unique(dft_stan$field_idx)),
  field = dft_stan$field_idx,
  y = dft_stan$disease_obs,
  Se_a = Se_beta$alpha,
  Se_b = Se_beta$beta,
  Sp_a = Sp_beta$alpha,
  Sp_b = Sp_beta$beta
)

# Model
mod <- cmdstanr::cmdstan_model("~/Desktop/WAVE_project/Theory/CassavaMap/misclass_hier_logit_nocov.stan")

fit <- mod$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 123
)

## Join with the posterieur

posterior_long <- as_draws_df(fit$draws()) %>%
  select(starts_with("p[")) %>%
  pivot_longer(everything(), names_to = "field_index", values_to = "p") %>%
  mutate(field_index = as.integer(gsub("p\\[|\\]", "", field_index))) %>%
  left_join(field_ids, by = c("field_index" = "field_idx")) %>%
  group_by(Unique_Field_ID, Commune, Departement, year) %>%
  summarise(
    true_prev_median = median(p) * 100,
    true_prev_lower = quantile(p, 0.025) * 100,
    true_prev_upper = quantile(p, 0.975) * 100,
    .groups = "drop"
  )




library(dplyr)
library(sf)
library(stringi)
library(ggplot2)
library(viridis)
library(ggspatial)
library(purrr)
library(akima)

# --- 1️⃣ Harmonize names in all data ----
posterior_spatial <- posterior_long %>%
  left_join(
    merged_filled_coords %>%
      select(Unique_Field_ID, Longitude, Latitude),
    by = "Unique_Field_ID"
  ) %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  mutate(
    Commune = stri_trans_general(Commune, "Latin-ASCII"),
    Commune = str_to_lower(str_trim(Commune)),
    Departement = stri_trans_general(Departement, "Latin-ASCII"),
    Departement = str_to_lower(str_trim(Departement))
  ) %>% 
  drop_na()

benin_lvl2 <- benin_lvl2 %>%
  mutate(
    Commune = stri_trans_general(Commune, "Latin-ASCII"),
    Commune = str_to_lower(str_trim(Commune)),
    Departement = stri_trans_general(Departement, "Latin-ASCII"),
    Departement = str_to_lower(str_trim(Departement))
  )

# --- 2️⃣ Compute commune-level posterior medians per year ----
commune_prev <- posterior_spatial %>%
  group_by(Departement, Commune, year) %>%
  summarise(
    commune_prev_median = median(true_prev_median, na.rm = TRUE),
    n_farms = n(),
    .groups = "drop"
  ) %>% 
  drop_na()

# --- 3️⃣ Join shapefile and posterior summaries ----
benin_map_all <- left_join(
  benin_lvl2,
  commune_prev,
  by = c("Commune", "Departement")
) %>% 
  filter(!is.na(year))

# --- 4️⃣ Create isocontours for each year ----
# Define interpolation and contouring function
make_isocontour <- function(data_year, benin_lvl2) {
  if (nrow(data_year) == 0) return(NULL)
  
  data_year <- data_year %>%
    mutate(
      Longitude_j = Longitude + runif(n(), 0, 1e-6),
      Latitude_j  = Latitude + runif(n(), 0, 1e-6)
    )
  
  interp_data <- akima::interp(
    x = data_year$Longitude_j,
    y = data_year$Latitude_j,
    z = data_year$true_prev_median / 100,
    duplicate = "mean",
    linear = TRUE,
    extrap = TRUE
  )
  
  z_min <- min(interp_data$z, na.rm = TRUE)
  z_max <- max(interp_data$z, na.rm = TRUE)
  thresholds <- seq(z_min + 1e-5, z_max, length.out = 5)
  
  iso <- isobands(
    x = interp_data$x,
    y = interp_data$y,
    z = interp_data$z,
    levels_low = thresholds[-length(thresholds)],
    levels_high = thresholds[-1]
  )
  
  iso_sfg <- iso %>%
    keep(~length(.x$x) > 0) %>%
    map(~tryCatch(iso_to_sfg(.x), error = function(e) NULL)) %>%
    compact() %>%
    flatten()
  
  if (length(iso_sfg) == 0) return(NULL)
  
  iso_sf <- st_sfc(iso_sfg, crs = st_crs(benin_lvl2)) %>%
    st_sf(level = thresholds[seq_along(iso_sfg)], geometry = .)
  
  iso_sf
}
library(isoband)

# Generate contours per year
iso_list <- posterior_spatial %>%
  split(.$year) %>%
  map(~make_isocontour(.x, benin_lvl2))

# --- 5️⃣ Combine and plot all years ----
ggplot() +
  geom_sf(data = benin_lvl2, fill = "grey90", color = "grey70", size = 0.2) +
  geom_sf(data = benin_map_all, aes(fill = commune_prev_median),
          color = "grey40", size = 0.25) +
  scale_fill_viridis_c(option = "magma", name = "Prevalence (%)",
                       na.value = "transparent") +
  geom_sf(data = do.call(rbind, iso_list), color = "red", fill = NA, size = 1) +
  facet_wrap(~year, ncol = 4) +
  annotation_scale(location = "bl", width_hint = 0.3, text_cex = 0.7,
                   bar_cols = c("grey20", "white")) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         height = unit(1, "cm"), width = unit(1, "cm"),
                         style = north_arrow_fancy_orienteering) +
  labs(
    title = "True Incidence Across Benin Communes (Posterior Median Estimates)",
    subtitle = "Non-sampled communes shown in grey transparency",
    x = "Longitude (°E)", y = "Latitude (°N)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    legend.position = "bottom",
    panel.grid = element_blank()
  )

#### Isocontour 

library(dplyr)
library(purrr)
library(akima)
library(isoband)
library(sf)
library(geosphere)
library(ggplot2)
library(viridis)
library(ggspatial)
library(stringi)

# --- 1️⃣ Prepare spatial posterior data ----
posterior_spatial <- posterior_long %>%
  left_join(
    merged_filled_coords %>% select(Unique_Field_ID, Longitude, Latitude),
    by = "Unique_Field_ID"
  ) %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  mutate(
    Commune = stri_trans_general(Commune, "Latin-ASCII"),
    Commune = str_to_lower(str_trim(Commune)),
    Departement = stri_trans_general(Departement, "Latin-ASCII"),
    Departement = str_to_lower(str_trim(Departement))
  ) %>%
  drop_na()

# Ensure both posterior_spatial and benin_lvl2 use the same CRS
posterior_spatial <- posterior_spatial %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)

if (st_crs(benin_lvl2) != st_crs(posterior_spatial)) {
  benin_lvl2 <- st_transform(benin_lvl2, st_crs(posterior_spatial))
}



benin_lvl2 <- st_transform(benin_lvl2, 4326)

posterior_spatial <- posterior_spatial %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(32631)

yearly_thresholds <- posterior_spatial %>%
  group_by(year) %>%
  summarise(thresh = quantile(true_prev_median, 0.85, na.rm = TRUE))

first_year <- min(posterior_spatial$year)
origin <- posterior_spatial %>%
  filter(year == first_year) %>%
  filter(true_prev_median == max(true_prev_median, na.rm = TRUE)) %>%
  dplyr::slice(1) %>%
  select(Longitude, Latitude)

Theta <- c(origin$Longitude, origin$Latitude)


make_isocontour_05 <- function(data_year, boundary_sf, threshold) {
  # Remove duplicates
  data_year <- data_year %>% distinct(Longitude, Latitude, .keep_all = TRUE)

  # Interpolate
  interp_data <- akima::interp(
    x = data_year$Longitude,
    y = data_year$Latitude,
    z = data_year$true_prev_median,
    duplicate = "mean"
  )

  # Extract contour lines at threshold
  contour_list <- contourLines(interp_data, levels = threshold)
  if (length(contour_list) == 0) return(NULL)

  # Convert to sf MULTILINESTRING object
  contour_sf <- map(contour_list, ~ st_linestring(cbind(.x$x, .x$y))) %>%
    st_sfc(crs = 4326) %>%       # make it a proper geometry collection
    st_sf(geometry = .)          # wrap as sf

  # Clip to Benin boundaries (ensure same CRS)
  if (st_crs(contour_sf) != st_crs(boundary_sf)) {
    boundary_sf <- st_transform(boundary_sf, st_crs(contour_sf))
  }

  contour_sf <- suppressWarnings(st_intersection(contour_sf, boundary_sf))

  return(contour_sf)
}


iso_list <- posterior_spatial %>%
  split(.$year) %>%
  imap(~ make_isocontour_05(
    .x, benin_lvl2,
    threshold = yearly_thresholds$thresh[
      yearly_thresholds$year == as.numeric(.y)
    ]
  )) %>%
  compact()  # remove NULLs


iso_all <- imap_dfr(iso_list, function(x, y) {
  if (inherits(x, "list")) x <- st_as_sf(do.call(rbind, x))  # flatten if needed
  if (!inherits(x, "sf")) return(NULL)
  x$year <- as.numeric(y)
  x
})
# Epi front
p1 <- ggplot() +
  geom_sf(data = benin_lvl2, fill = "grey95", color = "grey80") +
  geom_sf(data = iso_all, aes(color = year), fill = NA, size = 1.1) +
  geom_point(aes(x = Theta[1], y = Theta[2]), color = "black", size = 3) +
  scale_color_viridis_c(option = "plasma", direction = 1) +
  labs(
    title = "Temporal Progression of Infection Front (p = 0.5)",
    subtitle = sprintf("Origin Θ = (%.3f, %.3f)", Theta[1], Theta[2]),
    x = "Longitude", y = "Latitude", color = "Year"
  ) +
  theme_minimal()
print(p1)

# Direction of epidemics
front_positions <- iso_all %>%
  group_by(year) %>%
  summarise(center = st_centroid(st_union(geometry))) %>%
  arrange(year)

arrow_data <- st_coordinates(front_positions$center) %>%
  as.data.frame() %>%
  mutate(year = front_positions$year)

library(ggplot2)
library(ggspatial)
library(viridis)

library(ggplot2)
library(sf)
library(dplyr)
library(ggspatial)
library(viridis)

# Create one label position per department
benin_labels <- benin_lvl2 %>%
  group_by(Departement) %>%
  summarise(geometry = st_union(geometry)) %>%
  st_centroid()

# Shuffle department levels for more visual contrast between neighbors
benin_lvl2$Departement <- factor(benin_lvl2$Departement,
                                 levels = sample(unique(benin_lvl2$Departement)))

ggplot() +
  # Base map: departments with more distinct colors, transparent fills
  geom_sf(data = benin_lvl2,
          aes(fill = Departement),
          color = "grey50",
          alpha = 0.35) +

  # Isocontours by year (epidemic fronts)
  geom_sf(data = iso_all, aes(color = year), fill = NA, linewidth = 1.1) +

  # Arrow showing epidemic direction
  geom_path(data = arrow_data, aes(X, Y),
            color = "black",
            arrow = arrow(length = unit(0.3, "cm")),
            linewidth = 1) +

  # Origin point
  geom_point(aes(x = Theta[1], y = Theta[2]),
             color = "blue", size = 3) +

  # Department names at centroids
  geom_sf_text(data = benin_labels,
               aes(label = Departement),
               size = 3.5,
               color = "black",
               fontface = "italic",
               check_overlap = TRUE) +

  # Color scales
  scale_color_viridis_c(option = "plasma", direction = 1, name = "Year") +
  scale_fill_viridis_d(option = "turbo", guide = "none") +  # No legend for departments

  # Scale bar and north arrow
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering()) +

  # Labels
  labs(
    subtitle = sprintf("Origin Θ = (%.3f, %.3f)", Theta[1], Theta[2]),
    x = "Longitude",
    y = "Latitude"
  ) +

  # Theme
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "grey90", linetype = "dashed"),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.subtitle = element_text(size = 12)
  )


#-------Some statistics-------======
library(geosphere)  # for great-circle distance

# Origin coordinates
origin <- Theta  # e.g., c(2.5, 9.3)

# Ensure contours are valid SF geometries
iso_all <- st_make_valid(iso_all)

# If your contours are LINESTRINGs, convert to POLYGONs for area calculations
iso_poly <- st_cast(iso_all, "MULTILINESTRING", warn = FALSE)

# Compute centroids and areas
library(sf)
library(dplyr)

# Ensure geometries are in UTM
iso_poly_m <- st_transform(iso_poly, 32631) 
origin_m <- st_sfc(st_point(origin), crs = 4326) %>% st_transform(32631)

library(units)

iso_summary <- iso_poly_m %>%
  group_by(year) %>%
  summarise(
    threshold = 0.5,
    n_contours = sum(!st_is_empty(geometry)),
    
    mean_distance_km = {
      centroids <- st_centroid(geometry)
      centroids <- centroids[!st_is_empty(centroids)]
      if(length(centroids) == 0) NA
      else mean(st_distance(centroids, origin_m)) / 1000
    },
    
    max_distance_km = {
      centroids <- st_centroid(geometry)
      centroids <- centroids[!st_is_empty(centroids)]
      if(length(centroids) == 0) NA
      else max(st_distance(centroids, origin_m)) / 1000
    },
    
    area_km2 = {
      # buffer lines by small distance (e.g., 100 m) to get polygon
      polys <- st_buffer(geometry, dist = 100)  
      polys <- polys[!st_is_empty(polys)]
      if(length(polys) == 0) NA
      else sum(set_units(st_area(polys), km^2))
    },
    
    .groups = "drop"
  )


#===============
# Recompute the distance and estimate the uncertainties 

library(dplyr)
library(segmented)  # For piecewise regression

# --- Assume you have posterior samples in a list ---
# posterior_samples <- list(df1, df2, ...) 
# Each df has columns: field_id, t, prob_infected, x, y

posterior_samples <- posterior_spatial %>%
  st_drop_geometry() %>%  # remove geometry for computation
  split(.$year) 


calculate_front <- function(df, origin) {
  df %>%
    mutate(dist = sqrt((Longitude - origin[1])^2 + (Latitude - origin[2])^2)) %>%
    summarise(
      D_t = sum(dist * true_prev_median) / sum(true_prev_median),
      t = unique(year)
    )
}


# --- Step 1: Compute front positions for each posterior sample ---
front_positions_list <- lapply(posterior_samples, calculate_front, origin = Theta)

# Combine into a single dataframe for analysis
front_positions <- bind_rows(front_positions_list, .id = "sample_id")

# --- Step 2: Linear regression for each posterior sample ---
lm_fit <- lm(D_t ~ t, data = front_positions)
summary(lm_fit)
velocity_IC <- confint(lm_fit)["t", ]
#---------
# Add predictions and residuals
front_positions <- front_positions %>%
  mutate(pred = predict(lm_fit),
         resid = residuals(lm_fit))

# Plot actual vs fitted line
ggplot(front_positions, aes(x = t, y = D_t)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(aes(y = pred), color = "darkorange", linewidth = 1) +
  geom_segment(aes(xend = t, yend = pred), color = "gray60", linetype = "dashed") +
  geom_text(aes(label = round(resid, 2)), vjust = -1, color = "black", size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(x = "Year", y = "Mean front distance (km)",
       title = "BBTV front progression and linear fit",
       subtitle = "Dashed lines show residuals (deviation from linear trend)")





##======= Segmented regression
library(segmented)

# Try segmented regression with an initial guess near mid-time
try({
  seg_fit <- segmented(lm_fit, seg.Z = ~t, psi = list(t = median(front_positions$t)))
  summary(seg_fit)
  
  # Plot the fitted segmented model
  plot(front_positions$t, front_positions$D_t, pch = 16,
       xlab = "Year", ylab = "Mean front distance (km)",
       main = "Segmented regression: testing for change in spread velocity")
  plot(seg_fit, add = TRUE, col = "red", lwd = 2)
}, silent = FALSE)

ggplot(front_positions, aes(x = t, y = D_t)) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "loess", span = 0.8, se = TRUE, color = "red", linewidth = 1.2) +
  #geom_segment(aes(xend = t, yend = pred), color = "gray60", linetype = "dashed") +
  #geom_text(aes(label = round(resid, 2)), vjust = -1, color = "black", size = 3.5) +
  theme_minimal(base_size = 14) +
  labs(x = "Year", y = "Mean front distance (km)",
       title = "Smoothed BBTV front velocity trend",
       subtitle = "A visible slope increase around 2020–2022 indicates acceleration")



```



## Method in R:

Use spatial interpolation (e.g., kriging or inverse distance weighting) to map prevalence continuously across the landscape.

Identify the contour line corresponding to a low prevalence threshold (e.g., 5–10%) as the “front.”




```{r, echo=FALSE}

library(dplyr)
library(ggplot2)
library(sf)

# ---- 1. Compute mean vector from your gradient data ----
# grad_sub should contain columns x, y, dx, dy
dominant_vec <- grad_sub %>%
  summarise(dx = mean(dx, na.rm = TRUE),
            dy = mean(dy, na.rm = TRUE),
            x = mean(x, na.rm = TRUE),
            y = mean(y, na.rm = TRUE))

# ---- 2. Plot the map + IDW surface (transparent) ----
ggplot() +
  geom_sf(data = benin_lvl3 , fill = "white", color = "black") +
  geom_raster(data = r_pts, aes(x=x, y=y, fill=pred), alpha=0.6) +   # transparent IDW
  scale_fill_viridis_c(option="plasma", direction=-1, name="True Prevalence") +

  # ---- 3. Plot only the dominant direction ----
  geom_segment(data = dominant_vec,
               aes(x = x, y = y, xend = x + dx, yend = y + dy),
               arrow = arrow(length = unit(0.3, "cm")),
               size = 1.2, color = "black") +

  coord_sf() +
  theme_minimal() +
  labs(title = "Dominant Direction of Epidemic Spread",
       fill = "True Prevalence")


```


Is there any spatial cluster within the data?
